Estimating duration in the presence of misclassification

### 2023-09-12 incidence-prevalence.tex
Try and make this chapter reasonable...

From this, we can derive the relationship between incidence and prevalence.
Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
For each infection beginning at $t$, the probability that it ends at $t'$ is $f_d(t' - t + 1)$.
Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed.
Therefore:
\begin{align}
\begin{bmatrix}
  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
\end{bmatrix} \mid Z_t
\sim \MN \left(
  Z_t, 
  \begin{bmatrix}
    f_d(1) \\ f_d(2) \\ \vdots \\ f_d(d_\text{max})
  \end{bmatrix}
\right).
\end{align}
Now, denote by $P_{t,t'}$ the number of episodes beginning at time $t$ that are prevalent at time $t'$ (where $t' \geq t$).
This is $Z_t$ minus those that have recovered:
\begin{align}
    P_{t,t'} = \begin{cases}
      0 &t' < t\\
      Z_t - \sum_{i=0}^{t'-t-1} R_{t,t+i} &t \leq t' < t + d_\text{max}\\
      0 &t' > t + d_\text{max}.
  \end{cases}
\end{align}
Finally, the prevalence at $t$ can be expressed as $P_t = \sum_{t'=-\infty}^t P_{t',t}$.

@@

A quantity that will be important is $\E(P_t \mid \vec{Z})$, which can be derived from its constituent parts.
First, the properties of a multinomial distribution give the marginal distribution:
\begin{align}
  \sum_{i=t}^{t'-1} R_{t,i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, \sum_{i=1}^{t-t'} f_d(i)). \\
\intertext{Therefore:}
\E(P_{t-t'} \mid \vec{Z})
  &= \E(Z_t - \sum_{i=t}^{t'-1} R_{t,i} \mid \vec{Z}) \\
  &= Z_t (1 - \sum_{i=1}^{t-t'} f_d(i)).
\intertext{Finally:}
\E(P_t \mid \vec{Z})
&= \E(\sum_{t'=-\infty}^t P_{t-t'} \mid \vec{Z}) \\
&= \sum_{t'=-\infty}^t Z_t (1 - \sum_{i=1}^{t-t'} f_d(i)).
\end{align}
The quantity $1 - \sum_{i=1}^{t-t'} f_d(i)$ is the probability that an infection episode's duration is at least $t - t' + 1$, which is the survival function of $D$, denoted $S(t - t' + 1)$.
This equation, normally expressed in terms of the survival function, forms the basis of discrete-time \emph{backcalculation}.

@@

@@

A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
This thesis aims to infer the incidence rate from prevalence surveys.
I use Bayesian inference, therefore the data contributes through the likelihood.

The likelihood contribution from the survey at a time $t$ is based on a sample of $n_t$ individuals from the population of interest returning $x_t$ positive tests.
For now, I assume that the sample is chosen uniformly at random from the population of interest.
Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
The population prevalence is a latent quantity.
The likelihood is then $\p(\vec{x} \mid \vec{n}, \vec{Z}) = \int \p(\vec{P} \mid \vec{Z}) \prod_t \p(x_t \mid n_t, \vec{P}_t) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$ and similarly for $\vec{n}$.

In settings I consider, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
This can be seen by considering the variance in the likelihood.
$x_t$ is binomially distributed with parameters $n_t$ and $P_t / N$, where $N$ is the size of the population.
Therefore, by the law of total variance:
\begin{align}
  \var\left( x_t \mid \vec{n}, \vec{Z} \right)
    &= \var\left[\E\left( x_t \mid \vec{n}, \vec{Z}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] + \E\left[\var\left( x_t \mid \vec{n}, \vec{I}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] \\
    &= \var\left( n_t \frac{P_t}{N} \mid \vec{n}, \vec{Z} \right) + \E\left( n_t \frac{P_t}{N} \left(1 - \frac{P_t}{N} \right) \mid \vec{n}, \vec{Z} \right) \\
    &= \left( \frac{n}{N} \right)^2 \var\left(P_t \mid \vec{n}, \vec{Z} \right) + \frac{n_t}{N} \E\left(P_t \mid \vec{n}, \vec{Z} \right)  - \frac{n_t}{N^2} \E\left(P_t^2 \mid \vec{n}, \vec{Z} \right).
\end{align}
For large populations and reasonable sample sizes, $n_t << N$ and $\E(P_t) \approx \var(P_t)$, the first term is negligible.
Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, generally $\vec{P} = \E( \vec{P} \mid \vec{Z})$.

@@

\section{Estimating the duration}
\todo[inline]{Could be conclusion this section?}

@@

>>>>

From this, we can derive the relationship between incidence and prevalence.
Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
For each infection beginning at $t$, the probability that it ends at $t'$ is $f_d(t' - t + 1)$.
Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed.
Therefore:
\begin{align}
\begin{bmatrix}
  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
\end{bmatrix} \mid Z_t
\sim \MN \left(
  Z_t, 
  \begin{bmatrix}
    f_d(1) \\ f_d(2) \\ \vdots \\ f_d(d_\text{max})
  \end{bmatrix}
\right).
\end{align}
Now, denote by $P_{t_z,t_p}$ the number of episodes beginning at time $t_z$ that are prevalent at time $t_p$ (where $t' \geq t$).
This is $Z_t$ minus those that have recovered:
\begin{align}
    P_{t_z,t_p} = \begin{cases}
      0 &t_p < t_z\\
      Z_t - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
      0 &t_p \geq t_z + d_\text{max}.
  \end{cases} \label{incprev:eq:Ptt-to-R}
\end{align}
Finally, the prevalence at $t$ can be expressed as the total number of people prevalent at time $t$, regardeless of when they were infected:
\begin{align}
  P_t
  &= \sum_{t_z=\infty}^\infty P_{t_z,t} \\
  &= \sum_{i=0}^{d_\text{max}-1} P_{t-i,t} \label{incprev:eq:Pt-to-Ptt}.
\end{align}

@@

A quantity that will be important is $\E(P_t \mid \vec{Z})$, which can be derived from its constituent parts.
First, the properties of a multinomial distribution give the marginal distribution:
\begin{align}
  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_d(t'+1)).
\end{align}
where $F_d(t) = \sum_{i=1}^t f_d(t) = \prob(D \leq t)$ is the CDF of $D$.
Therefore:
\begin{align}
  \E(\sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z}) = Z_t F_d(t'+1) \label{incprev:eq:EsumRt}
\end{align}

Now consider $P_t$:
\begin{align}
\E(P_t \mid \vec{Z})
  &= \E\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Pt-to-Ptt}} \\
  &= \E\left(\sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Ptt-to-R}}\\
  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F(i) \right) &\text{by \cref{incprev:eq:EsumRt}}\\
  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F(i)) \\
  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} S(i+1)
\end{align}
where $S(i+1) = 1-F(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{inprev:sec:duration}).

@@

\todo[inline]{Need some conclusion text for this section}

@@

% A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
% Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
% This thesis aims to infer the incidence rate from prevalence surveys.
% I use Bayesian inference, therefore the data contributes through the likelihood.

Prevalence surveys, as described in\todo{insert ref}, sample individuals at random from the population of interest and test if they are prevalence.
Therefore, we view the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}_t$. 
%For now, I assume that the sample is chosen uniformly at random from the population of interest.
%Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
%The population prevalence is a latent quantity.
From these assumptions, $X_t \mid \vec{P} \sim \Bin(n_t, P_t / N)$.
I base inference on the likelihood, $p(\vec{x} \mid \vec{Z}) = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.

For most prevalence surveys, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
This can be seen by considering the variance in the likelihood.
Assume $x_t \sim \Bin(n_t, P_t / N)$, where $N$ is the size of the population.
Therefore, by the law of total variance:
\begin{align}
  \var\left( x_t \mid \vec{n}, \vec{Z} \right)
    &= \var\left[\E\left( x_t \mid \vec{n}, \vec{Z}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] + \E\left[\var\left( x_t \mid \vec{n}, \vec{I}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] \\
    &= \var\left( n_t \frac{P_t}{N} \mid \vec{n}, \vec{Z} \right) + \E\left( n_t \frac{P_t}{N} \left(1 - \frac{P_t}{N} \right) \mid \vec{n}, \vec{Z} \right) \\
    &= \left( \frac{n}{N} \right)^2 \var\left(P_t \mid \vec{n}, \vec{Z} \right) + \frac{n_t}{N} \E\left(P_t \mid \vec{n}, \vec{Z} \right)  - \frac{n_t}{N^2} \E\left(P_t^2 \mid \vec{n}, \vec{Z} \right).
\end{align}
For large populations and reasonable sample sizes, $n_t << N$ and $\E(P_t) \approx \var(P_t)$, the first term is negligible.
Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, $\vec{P} = \E( \vec{P} \mid \vec{Z})$.

@@

\todo[inline]{Explain backcalculation in HIV/AIDS context, and how it relates to what is above.}

\section{Estimating the duration} \label{inprev:sec:duration}

@@

\todo[inline]{Comment on previous estimates of duration for SARS-CoV-2, but not sure where exactly this goes...}
